---
title: "Exploring Cross-country Literacy Statistics"
author: "Quynh (Christina) Vu"
date: ""
output: 
  html_document:
    theme: spacelab
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library (readr)
library(tidyverse)
library(brms)
library(tidybayes)
library(rvest)
library(tidyr)
library(gtable)
library(grid)
library(gridExtra)
`%nin%` <- negate(`%in%`) # this will later help  
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
literacy_data <- read.csv("https://raw.githubusercontent.com/pquynhvu/bayes/refs/heads/main/global-literacy-rate/cross-country-literacy-rates.csv") |> 
                 rename("country" = "Entity", 
                        "ISO_code" = "Code", 
                        "year" = "Year", 
                        "literacy_rate" = "Literacy.rate") |>
                 group_by(country) |> mutate(n_obs = n()) |> ungroup() |>
                 mutate(across(where(is.character), ~ na_if(.,""))) |> # replace empty ISO_code w/ NA
                 filter(!is.na(ISO_code)) # remove "inappropriate" countries
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
## Match countries with regions & continents by ISO code
ISO_page <- read_html("https://cloford.com/resources/codes/index.htm")
ISO_code <- ISO_page %>% html_nodes("p+ .outlinetable td:nth-child(7)") %>% html_text() 
region <- ISO_page %>% html_nodes("p+ .outlinetable tr :nth-child(2)") %>% html_text()
region <- region[-1]
continent <- ISO_page %>% html_nodes("p+ .outlinetable td:nth-child(1)") %>% html_text() 
ISO_data <- data.frame(ISO_code, region, continent)
```


```{r, echo=FALSE, warning=FALSE, message=FALSE}
literacy_data_cleaned <- left_join(literacy_data, ISO_data, by = "ISO_code") |>
                         filter(country %nin% c("World", "Saint Helena", "French Polynesia", "Gibraltar",
                                                "Hong Kong", "Macao", "Anguilla", "American Samoa",
                                                "Bermuda", "British Virgin Islands", "Montserrat",
                                                "Turks and Caicos Islands", "Puerto Rico", "Guam",
                                                "Cayman Islands", "Cook Islands", "Aruba",
                                                "Northern Mariana Islands", "New Caledonia",
                                                "Saint Pierre and Miquelon", "Wallis and Futuna")) |> 
                         mutate(region = ifelse(country == "East Timor", replace_na(region, "South East Asia"), region),
                                continent = ifelse(country == "East Timor", replace_na(continent, "Asia"), continent),
                                region = ifelse(country == "Palestine", replace_na(region, "South West Asia"), region),
                                continent = ifelse(country == "Palestine", replace_na(continent, "Asia"), continent),
                                region = ifelse(country == "Kosovo", replace_na(region, "South East Europe"), region),
                                continent = ifelse(country == "Kosovo", replace_na(continent, "Europe"), continent),
                                region = ifelse(country == "Montenegro", replace_na(region, "South East Europe"), region),
                                continent = ifelse(country == "Montenegro", replace_na(continent, "Europe"), continent),
                                region = ifelse(country == "South Sudan", replace_na(region, "Eastern Africa"), region),
                                continent = ifelse(country == "South Sudan", replace_na(continent, "Africa"), continent),
                                region = ifelse(country == "Zambia", replace_na(region, "Southern Africa"), region),
                                continent = ifelse(country == "Zambia", replace_na(continent, "Africa"), continent),
                                region = ifelse(country == "Romania", replace_na(region, "South East Europe"), region),
                                continent = ifelse(country == "Romania", replace_na(continent, "Europe"), continent),
                                region = ifelse(country == "Serbia", replace_na(region, "South East Europe"), region),
                                continent = ifelse(country == "Serbia", replace_na(continent, "Europe"), continent),
                                region = ifelse(country == "Democratic Republic of Congo", replace_na(region, "Central Africa"), region),
                                continent = ifelse(country == "Democratic Republic of Congo", replace_na(continent, "Africa"), continent)) |>
                        mutate(region = ifelse(region == "West Indies", "Caribbean", as.character(region))) |>
                        arrange(country)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
country_fill <- literacy_data_cleaned |> select(country, region, continent) |> 
                                         filter(if_any(everything(), is.na)) |> 
                                         unique() |>
                                         select(- c(region, continent)) 
```

# Introduction

Being able to read and write opens up the world of education and knowledge. When and why did more people become literate? How can progress continue?

Only consider countries individually and not territory (a, 

British Overseas Territories: Saint Helen, Anguilla, Bermuda, British Virgin Islands, Turks and Caicos Islands, Montserrat, Cayman Islands, Gibraltar


US terr:  American Samoa, Puerto Rico, 	Guam, Northern Mariana Islands, 

China special administative: Hong Kong, Macao

New Zealan: Cook Islands

Kingdom of the Netherlands: 	 Aruba

France: French Polynesia, New Caledonia, Saint Pierre and Miquelon, Wallis and Futuna

```{r, echo=FALSE, warning=FALSE, message=FALSE}
literacy_data_latest <- literacy_data_cleaned |> group_by(ISO_code) |> slice(which.max(year)) |> ungroup() 
```


```{r, echo=FALSE, warning=FALSE, message=FALSE}
plot1 <- literacy_data_latest |> group_by(continent) |>
                                 summarise(avg_literacy_rate = mean(literacy_rate)) |>
                                 ggplot(aes(x=reorder(continent, avg_literacy_rate), y=avg_literacy_rate)) +
                                        geom_bar(stat = "identity", , fill="brown1", width=0.5) +
                                  geom_hline(yintercept = mean(literacy_data$literacy_rate)) +
                                  annotate("text", x=1.1, y=85, label= "Global average literacy rate") + 
                                  labs(title="", subtitle="", x="continents", y="Average literacy rates") +
                                  theme_minimal()
plot2 <- literacy_data_latest |> ggplot(aes(year, literacy_rate)) + 
                                        geom_line(aes(linetype = continent))

plot2

grid.arrange(plot1, plot2,nrow=2)
```

Data source: [United Nations Inter-agency Group for Child Mortality Estimation (2024)](https://ourworldindata.org/literacy) 

Demographic estimation usually involves using some sort of model process or set of assumptions, no matter