---
title: "Exploring Cross-country Literacy Statistics"
author: "Quynh (Christina) Vu"
date: ""
output: 
  html_document:
    theme: spacelab
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library (readr)
library(tidyverse)
library(brms)
library(tidybayes)
library(rvest)
library(tidyr)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
literacy_data <- read.csv("https://raw.githubusercontent.com/pquynhvu/bayes/refs/heads/main/global-literacy-rate/cross-country-literacy-rates.csv")
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
`%nin%` <- negate(`%in%`) # this will later help  
```

# Introduction

Being able to read and write opens up the world of education and knowledge. When and why did more people become literate? How can progress continue?

Only consider countries individually and not territory (American Samoa, British Virgin Islands, Turks and Caicos Islands, 	
Anguilla, Virgin Islands) using latest data available "Bermuda", "Cayman Islands", French Polynesia, Wallis and Futuna

Saint Pierre and Miquelon

Saint Helena

Northern Mariana Islands

Montserrat

Macao

Guam

Gibraltar

Hong Kong 

```{r, echo=FALSE, warning=FALSE, message=FALSE}
literacy_data <- literacy_data %>% rename("country" = "Entity", 
                                          "ISO_code" = "Code", 
                                          "year" = "Year", 
                                          "literacy_rate" = "Literacy.rate") 
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
## Match countries with regions & continents by ISO code
ISO_page <- read_html("https://cloford.com/resources/codes/index.htm")
ISO_code <- ISO_page %>% html_nodes("p+ .outlinetable td:nth-child(7)") %>% html_text() 
region <- ISO_page %>% html_nodes("p+ .outlinetable tr :nth-child(2)") %>% html_text()
region <- region[-1]
continent <- ISO_page %>% html_nodes("p+ .outlinetable td:nth-child(1)") %>% html_text() 
ISO_data <- data.frame(ISO_code, region, continent)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
literacy_data <- left_join(literacy_data, ISO_data, by = "ISO_code") %>%
                 filter(country %nin% c('World', 'Upper middle income', 'Anguilla', "Bermuda", "Cayman Islands",
                                        'Upper-middle-income countries','Early-demographic dividend', 
                                        'Western and Central Africa (WB)', 'Sub-Saharan Africa', "Wallis and Futuna",
                                        'Sub-Saharan Africa (IDA & IBRD)', 'Sub-Saharan Africa (WB)', 
                                        'Sub-Saharan Africa (excluding high income)', 'Caribbean small states',
                                        'Central Europe and the Baltics', 'Central Europe and the Baltics (WB)', 
                                        'East Asia & Pacific', 'East Asia & Pacific (IDA & IBRD)', "Saint Helena",
                                        'East Asia & Pacific (excluding high income)', 'East Asia and the Pacific (WB)', 
                                        'Europe & Central Asia', 'Europe & Central Asia (IDA & IBRD)',         
                                        'Europe & Central Asia (excluding high income)', 'Europe and Central Asia (WB)',
                                        'Fragile and conflict affected situations', 'Heavily indebted poor countries (HIPC)',
                                        'IBRD only', 'IDA & IBRD total', 'IDA blend', 'IDA only', 'IDA total', 
                                        'Late-demographic dividend', 'Latin America & Caribbean', "Democratic Republic of Congo",
                                        'Latin America & Caribbean (IDA & IBRD)', "American Samoa", 
                                        'Latin America & Caribbean (excluding high income)', "British Virgin Islands",
                                        'Latin America and Caribbean (WB)', 'Low income',  'Low-income countries', 
                                        'Lower middle income', 'Lower-middle-income countries', 'Middle East & North Africa',
                                        'Middle East & North Africa (IDA & IBRD)', "Arab World (WB)", "Arab World",
                                        'Middle East & North Africa (excluding high income)', "Saint Pierre and Miquelon",
                                        'Middle East and North Africa (WB)', 'Middle income', 'North America (WB)', 
                                        'Other small states', 'Pacific island small states', 'Pre-demographic dividend', 
                                        'Small states', 'South Asia', 'South Asia (IDA & IBRD)', 'South Asia (WB)', 
                                        'Southern and Eastern Africa (WB)', 'Least developed countries: UN classification',
                                        "Turks and Caicos Islands", "Micronesia", "Virgin Islands", "French Polynesia",
                                        "Northern Mariana Islands", "Montserrat", "Macao", "Guam", "Gibraltar", "Hong Kong")) %>% 
                 group_by(country) %>%
                 slice(which.max(year)) %>% 
                 ungroup()
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
country_fill <- literacy_data %>% select(country, region, continent) %>% 
                                  filter(if_any(everything(), is.na)) %>% 
                                  unique() %>% 
                                  select(- c(region, continent))
                                  
literacy_data <- literacy_data %>% mutate(region = ifelse(country == "East Timor", replace_na(region, "South East Asia"), region),
                                          continent = ifelse(country == "East Timor", replace_na(continent, "Asia"), continent),
                                          region = ifelse(country == "Kosovo", replace_na(region, "South East Europe"), region),
                                          continent = ifelse(country == "Kosovo", replace_na(continent, "Europe"), continent),
                                          region = ifelse(country == "Montenegro", replace_na(region, "South East Europe"), region),
                                          continent = ifelse(country == "Montenegro", replace_na(continent, "Europe"), continent),
                                          region = ifelse(country == "Palestine", replace_na(region, "South West Asia"), region),
                                          continent = ifelse(country == "Palestine", replace_na(continent, "Asia"), continent),
                                          region = ifelse(country == "Romania", replace_na(region, "South East Europe"), region),
                                          continent = ifelse(country == "Romania", replace_na(continent, "Europe"), continent),
                                          region = ifelse(country == "Serbia", replace_na(region, "South East Europe"), region),
                                          continent = ifelse(country == "Serbia", replace_na(continent, "Europe"), continent),
                                          region = ifelse(country == "South Sudan", replace_na(region, "Eastern Africa"), region),
                                          continent = ifelse(country == "South Sudan", replace_na(continent, "Africa"), continent),
                                          region = ifelse(country == "Zambia", replace_na(region, "Southern Africa"), region),
                                          continent = ifelse(country == "Zambia", replace_na(continent, "Africa"), continent))
```


```{r, echo=FALSE, warning=FALSE, message=FALSE}
plot1 <- literacy_data %>% group_by(continent) %>% 
                           summarise(avg_literacy_rate = mean(literacy_rate)) %>%
                           ggplot(aes(x=reorder(continent, avg_literacy_rate), y=avg_literacy_rate)) +
                                  geom_bar(stat = "identity", width=0.5, fill=c("chocolate1", "chartreuse", "blue", "gold3", 
                                                                                "firebrick1")) +
                                  geom_hline(yintercept = mean(literacy_data$literacy_rate)) +
                                  annotate("text", x=1.15, y=95, label= "Global average literacy rate") + 
                                  labs(title="", subtitle="", x="continents", y="Average literacy rates") +
                                  theme_minimal()
plot1
```

Data source: [United Nations Inter-agency Group for Child Mortality Estimation (2024)](https://ourworldindata.org/literacy) 

Demographic estimation usually involves using some sort of model process or set of assumptions, no matter